---
title: "(Part 2.2) United States Anti-Doping Agency Sanction Data - S1 Anabolic Steroids"
author: "Martin Frigaard"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
library(hrbrthemes)
library(extrafont)
library(knitr)
library(rmdformats)
library(knitr)
library(blogdown)
library(tidyverse)
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  prompt = FALSE,
  tidy = FALSE,
  fig.width = 10,
  fig.height = 6.5,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.path = "figs/"
)
# set theme
ggplot2::theme_set(hrbrthemes::theme_ipsum_tw(
  base_size = 8,
  strip_text_size = 9,
  axis_title_size = 9,
  plot_title_size = 15,
  subtitle_size = 13,
  base_family = "Ubuntu",
  strip_text_family = "TitilliumWeb-Regular",
  axis_title_family = "TitilliumWeb-Regular",
  subtitle_family = "TitilliumWeb-Regular",
  plot_title_family = "JosefinSans-Regular"
))
```

**Packages**

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)
library(skimr)
library(rlang)
library(gt)
library(visdat)
library(inspectdf)
library(foreign)
library(naniar)
```

# Motivation 

In these scripts (`02.2`), we will be wrangling the substances from the USADA website. 

## Import data 

In the previous post, I wrangled the data into a new data structure that looks different from the original data on the [USADA website](https://www.usada.org/testing/results/sanctions/). These data are located in the `data/proessed` folder. I can use the `fs::dir_ls()` functions `regexp` argument to find the most recent files. 

```{r usada_sanct_csv_data}
# check
ProcessedData <- fs::dir_ls("data/processed", 
      regexp = base::noquote(lubridate::today())) %>% 
      as_tibble()
ProcessedData
```

Here we can import the latest sanctions data set. 

```{r import-processed_data}
# choose the csv
usada_sanct_csv_data <- ProcessedData %>% 
    dplyr::filter(stringr::str_detect(string = value, 
          pattern = "UsadaSanctions")) %>% 
    as_vector() %>% 
    base::unname()

# choose the RData
usada_nonanalytic_rdata <- ProcessedData %>% 
    dplyr::filter(stringr::str_detect(string = value, 
          pattern = "02.1-usada-non-analytic")) %>% 
    as_vector() %>% 
    base::unname()

# import UsadaSanctions
UsadaSanctions <- readr::read_csv(file = usada_sanct_csv_data)
```

```{r}
# import 02.0-usada-sanction-dates.RData
base::load(ProcessedData$value[3])
ls()
```

Below is the `UsadaSanctions` data frame. This data set has the original dimensions (5 columns, 571 observations). 

```{r UsadaSanctions}
UsadaSanctions %>% dplyr::glimpse(78)
```


## The World Anti-Doping Agency’s (WADA) Prohibited List

USADA follows the list of prohibited substances published by the The World Anti-Doping Agency (WADA). The first list of prohibited substances was printed by the International Olympic Committee (IOC) in 1967. In the 1972 Munich Games, athletic testing was introduced. The current list of banned substances can be found on [WADA's website](https://www.usada.org/substances/prohibited-list/). A substance ends up banned when it meets two out of three of the following criteria:

1. It has the potential to enhance sport performance      
2. It represents an actual or potential health risk to the athlete     
3. It violates the spirit of sport   

These are all stored in the `substance_reason` variable, but I want them in the `substances` vector.

```{r substances}
substances <- sort(unique(UsadaSanctions$substance_reason))
substances %>% head()
```

We're going to go over the first agent, anabolic steroids.

### * S1 Anabolic agents

These are typically what people think of when they hear the word, "steroids." **Anabolic agents** (and their synthetic derivatives)  typically refer to [testosterone](https://en.wikipedia.org/wiki/Testosterone), a sex hormone produced by our body under normal conditions (although in slightly higher concentrations in males compared to females). 

Hormones are made, stored, and released by our endocrine glands. They work like chemical messengers, communicating important information to various body systems. This includes messages like building up ("anabolic") or breaking down ("catabolic") muscle cells and tissue. Testosterone and it's precursors are the most commonly used anabolic agents because of their role in building and maintaining skeletal muscle tissue.   

Perhaps the most famous example of anabolic steroid use was by sprinter [Ben Johnson](https://www.theguardian.com/australia-news/2017/jun/06/sportsbets-ben-johnson-ad-banned-for-making-light-of-drug-use) and the other athletes in the [1988 Summer Olympics](https://en.wikipedia.org/wiki/Athletics_at_the_1988_Summer_Olympics_%E2%80%93_Men%27s_100_metres). Johnson took [Stanozolol](https://en.wikipedia.org/wiki/Stanozolol), an anabolic agent popular among sprinters because of it's ability to gain lean muscle mass.  

**Steroid Cycles:** Generally speaking, anabolic agents are used in the pre-competition or "building" phase to enhance an athlete's training, with a focus on developing size, power, or strength. Starting and stopping steroids use throughout a competitive season is referred to as "cycling" (or a "cycle"). The duration of cycles vary, but generally don't exceed 8 weeks. Sports and athletic events in which size or strength contribute to performance have a potential market for anabolic agents. In fact, coaches and athletes use various resistance training programs and diets to try and stimulate these processes throughout an athletic career. 

### * S2 Peptide hormones, growth factors, related substances, and mimetics

**What causes muscle growth?** Training causes tiny tears and injuries to muscle tissues. These tiny injuries initiate a cascade of chemical messages that fills the muscle cells with more contractile proteins (actin and myosin). Collectively, this process is referred to as [hypertrophy](https://en.wikipedia.org/wiki/Hypertrophy). It results in an increase in muscle size and protein content, but it does not typically result in the production of more muscle cells. [Hyperplasia](https://en.wikipedia.org/wiki/Hyperplasia) is the creation of new muscle cells, and its still unclear if prolonged resistance training can stimulate this response.  

We've established earlier that hormones carry specific chemical messages to various cells throughout the body. Specific messages are intended for specific hormone receptors, like a key made to fit in a lock. These receptors vary by location (the cell membrane vs. inside the cell) and type (peptide vs. steroid hormone receptors). Increasing the amount of anabolic hormone messages can stimulate muscle growth, but so can increasing the number of receptors for that hormone. For example, if I flood my bloodstream with testosterone, and very few receptors are around to hear that "build muscle" message (and communicate it to my muscle tissue cells), nothing would change. However, if I increased the rate this specific signal was heard by increasing the number of receptors, I could indirectly boost the strength of that chemical message. 

Peptide hormones are also produced by the endocrine system and used by athletes to stimulate muscle growth and gains in strength, speed, or power.

[Insulin](https://en.wikipedia.org/wiki/Insulin), a peptide hormone, is often used in combination with testosterone. Insulin sends a chemical signal that stimulates the absorption of nutrients into the muscle cells. When used with an anabolic agent like testosterone, an athlete can stay in a "sustained anabolic state."

[Growth hormone](https://en.wikipedia.org/wiki/Growth_hormone) (GH) is produced by the pituitary gland and, as the name implies, causes growth in just about every human tissue. Bones, muscles, and even internal organs respond to GH. Our bodies produce higher levels of this hormone as we develop throughout childhood, but it's use and availability diminishes as we age. Animal models have shown growth hormone causes increases in both the size (hypertrophy) and number (hyperplasia) of muscle cells. A [study](https://www.ncbi.nlm.nih.gov/pubmed/18504393) in 2008 showed that giving adult males growth hormone resulted in increases in strength and power.  

These are stored in the string `anabolic_agents` below

```{r anabolic_agents}
anabolic_agents <- c(
  "1-androstenedione",

  "1-testosterone and 1-androstenedione",

  "17β-hydroxymethyl-17α-methyl-18-norandrost-1,4,13-trien-3-one",

  "19-norandrostendione",
  "19-norandrosterone",
  "19-norandrosterone (19-na)",
  "19-norandrosterone & 19-noretiocholanolone",
  "19-norandrosterone and 19-noretiocholanolone",
  "19-norandrosterone and testosterone",
  "19-norandrosterone, epo, furosemide",

  "6a-hydroxyandrostendione",
  "6a-oh-androstenedione & 6b-ohandrosterone",
  "anastrozole",
  
  "androgenic-anabolic steroids",
  "androgenic anabolic steroid",
  "androstendione",
  "androstenedione",
  "androstenedione, ostarine",
  "dehydrochloromethyltestosterone (dhcmt)",
  "dehydroepiandrosterone (“dhea”)",
  "dehydroepiandrosterone (dhea)",

  "boldenone",
  "boldenone and 19-norandrosterone",
  "boldenone metabolite",
  "boldione",
  "carboxy finasteride",
  "dianabol",

  "human growth hormone",
  "human growth hormone (hgh)",
  "human growth hormone (hgh) and testosterone",

  "clenbuterol",
  "clenbuterol, ostarine, and a tamoxifen metabolite",
  "clomiphene",
  "clomiphene and gw1516",
  "clomiphene, letrozole",
  "clomiphene, methylphenidate metabolite",
  "formestane (4-hydroxyandrostendione)",

  "testosterone",
  "testosterone & modafinil",
  "testosterone & recombinant human erythropoietin (r-epo)",
  "testosterone and boldenone",
  "testosterone metabolites (exogenous origin)",
  "testosterone prohormones",
  "testosterone, methandienone (dianabol) and stanzolol",
  "testosterone, nandrolone & hcg",
  "dexamethasone",
  "drostanolone",
  "elevated testosterone-epitestosterone (t/e) ratio",
  "exogenous testosterone",
  "tetrahydrogestrinone (thg)",
  "finasteride",
  "finasteride & 6-oxo-androstenedione",

  "anabolic agent",
  "anabolic agents",

  "ghrp-2",
  "growth hormone-releasing hexapeptide",
  "gw1516",
  "hormone antagonists, hormones and related substance, hcg, and an anabolic agent",
  "insulin-like growth factor-1",
  "methandienone",
  "methandienone (dianabol)",
  "ibutamoren",
  "meldonium",
  "metabolites of stanozolol",
  "methasterone, ostarine, androstatrienedione",
  "methenolone, androst-(2,3)-en-17-one, and ghrp-6",
  "methylclostebol, promagnon",
  "methylprednisolone",
  "methyltestosterone",
  "synthetic testosterone",
  "proplyhexedrine; anabolic steroid",
  "nandrolone metabolites",
  "norandrosterone",
  "stanozolol",
  "stanozolol / methylphenidate",
  "stanozolol metabolites",
  "stanzolol",
  "ostarine",
  "ostarine and 7-keto-dhea",
  "ostarine, arimistane",
  "oxandrolone",
  "oxandrolone and clenbuterol",
  "oxymetholone",
  "prednisone",
  "rad140 and ibutamoren",
  "tamoxifen",
  "te ratio > 4:1 & hcg",
  "trenbolone",
  "trenbolone, epitrenbolone",
  "triamcinolone acetonide",
  "ostarine"
)
anabolic_agents_regex <- paste0("(", stringr::str_c(anabolic_agents, 
                                                    collapse = "|"), ")")
```

```{r str_view_all-anabolic_agents_regex}
stringr::str_view_all(string = substances, 
                      pattern = anabolic_agents_regex, 
                      match = TRUE)
```

This looks like this is matching all the terms, so I can assign this to the data frame using a little trick with `dplyr::mutate()` and `dplyr::case_when()`. The regex pattern can be used in combination with `stringr::str_detect()`

```{r UsadaAnabolic-wada_cat}
UsadaAnabolic <- UsadaSanctions %>%
  dplyr::select(athlete, 
                sport, 
                substance_reason) %>% 
  dplyr::mutate(
      # create wada category
    wada_cat =
      case_when(
        # # non-analytic first
        stringr::str_detect(string = substance_reason,
                       pattern = "non-analytic") ~ "non-analytic",
        # now steroids
        stringr::str_detect(string = substance_reason, 
                        pattern = anabolic_agents_regex) ~ "S1 Anabolic Agents",
        TRUE ~ NA_character_))
UsadaAnabolic %>% dplyr::glimpse(78)
```

We will focus on these anabolic agents to make sure they're appropriately categorized. We will look at the anabolic steroids that show up more than once. 

```{r anabolic_agents-top-5}
ggp_anabolic_top5 <- UsadaAnabolic %>%
    dplyr::count(substance_reason, sort = TRUE) %>% 
    dplyr::mutate(substance_reason = reorder(substance_reason, n),
                  anabolic_agent_count = n) %>% 
    dplyr::filter(stringr::str_detect(
        string = substance_reason, pattern = anabolic_agents_regex),
        anabolic_agent_count > 5) %>% 
    ggplot2::ggplot(aes(substance_reason, anabolic_agent_count)) + 
    ggplot2::geom_col() + 
    ggplot2::labs(x = "Substance or reason for sanction", 
              y = "Anabolic agent count",
              title = "Anabolic Agent Sanctions",
              subtitle = "substances that occur 5+ times") + 
    ggplot2::coord_flip()
ggp_anabolic_top5
```

Now we will look at these by anabolic agents that occurred less frequently.

```{r anabolic_agents-less-than-5}
ggp_anabolic_low5 <- UsadaAnabolic %>%
    dplyr::count(substance_reason, sort = TRUE) %>% 
    dplyr::mutate(substance_reason = reorder(substance_reason, n),
                  anabolic_agent_count = n) %>% 
    dplyr::filter(stringr::str_detect(
        string = substance_reason, pattern = anabolic_agents_regex),
        anabolic_agent_count <= 5) %>% 
    ggplot2::ggplot(aes(substance_reason, anabolic_agent_count)) + 
    ggplot2::geom_col() + 
    ggplot2::labs(x = "Substance or reason for sanction", 
          y = "Anabolic agent count",
          title = "Less Occurring Anabolic Agent Sanctions",
          subtitle = "substances that occur less than 5 times") +
    ggplot2::coord_flip()
ggp_anabolic_low5
```

Now we will check the two categories in the `wada_cat` variable. 

```{r ggp_anabolic_cat}
ggp_anabolic_cat <- UsadaAnabolic %>% 
    dplyr::count(substance_reason, wada_cat, sort = TRUE) %>% 
    dplyr::ungroup() %>% 
    dplyr::group_by(wada_cat) %>% 
    dplyr::top_n(10) %>%
    dplyr::ungroup() %>% 
    dplyr::mutate(substance_reason = reorder(substance_reason, n)) %>% 
    ggplot2::ggplot(aes(substance_reason, n, fill = wada_cat)) + 
    ggplot2::geom_col(show.legend = FALSE) + 
    ggplot2::facet_wrap(~wada_cat, 
                 scales = "free_y", 
                 nrow = 3,
                 ncol = 2) + 
    ggplot2::labs(y = "No of sanctions/substances", 
                  x = NULL, 
                  title = "WADA list categories for banned substances",
                  subtitle = "Top ten substances/reason in each category") + 
    ggplot2::coord_flip()
ggp_anabolic_cat
```

We can see the `S1 Anabolic Agents` are clearly labeled, and the most common substance is the `androgenic anabolic steroid`. But there are still quite a few in the `NA` category, specifically `cannabinoids`.

## Export data 

We'll export this image and move onto the S3 Beta-2 agonists.

```{r processed-data}
base::save.image(file = paste0("data/processed/",
                               base::noquote(lubridate::today()), 
                               "-02.2-usada-S1-anabolic-steroids.RData"))
fs::dir_tree(path = "data/processed", 
           regexp = base::noquote(lubridate::today()))
```