---
title: "(Part 4.1) United States Anti-Doping Agency Sanction Data - Substances"
author: "Martin Frigaard"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
library(hrbrthemes)
library(extrafont)
library(knitr)
library(rmdformats)
library(knitr)
library(blogdown)
library(markdowntemplates)
library(tidyverse)
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  prompt = FALSE,
  tidy = FALSE,
  fig.width = 10,
  fig.height = 7,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.path = "figs/"
)
# set theme
ggplot2::theme_set(theme_ipsum_tw(
  base_size = 9,
  strip_text_size = 9,
  axis_title_size = 11,
  plot_title_size = 17,
  subtitle_size = 13,
  base_family = "Ubuntu",
  strip_text_family = "TitilliumWeb-Regular",
  axis_title_family = "TitilliumWeb-Regular",
  subtitle_family = "TitilliumWeb-Regular",
  plot_title_family = "JosefinSans-Regular"
))
```

**Packages**

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)
library(skimr)
library(rlang)
library(gt)
library(visdat)
library(inspectdf)
library(foreign)
library(naniar)
```

# Motivation 

This is the third in the series of posts on the United States Anti-Doping Agency sanction data project (see [post 1]() and [post 2]()).

## Import data 

In the previous post, I wrangled the data into a new data structure that looks different from the original data on the [USADA website](https://www.usada.org/testing/results/sanctions/). These data are located in the `data/proessed` folder. I can use the `fs::dir_ls()` functions `regexp` argument to find the most recent files. 

```{r usada_sanct_csv_data}
# check
processed_data <- fs::dir_ls("data/processed", 
      regexp = base::noquote(lubridate::today()))
# choose the csv
usada_sanct_csv_data <- processed_data[2]
# import
UsadaSanctions <- readr::read_csv(file = usada_sanct_csv_data)
```

Below is the `UsadaSanctions` data frame. This data set has the original dimensions (5 columns, 571 observations). 

```{r UsadaSanctions}
UsadaSanctions %>% dplyr::glimpse(78)
```


### * S1 Anabolic agents

These are typically what people think of when they hear the word, "steroids." **Anabolic agents** (and their synthetic derivatives)  typically refer to [testosterone](https://en.wikipedia.org/wiki/Testosterone), a sex hormone produced by our body under normal conditions (although in slightly higher concentrations in males compared to females). 

Hormones are made, stored, and released by our endocrine glands. They work like chemical messengers, communicating important information to various body systems. This includes messages like building up ("anabolic") or breaking down ("catabolic") muscle cells and tissue. Testosterone and it's precursors are the most commonly used anabolic agents because of their role in building and maintaining skeletal muscle tissue.   

Perhaps the most famous example of anabolic steroid use was by sprinter [Ben Johnson](https://www.theguardian.com/australia-news/2017/jun/06/sportsbets-ben-johnson-ad-banned-for-making-light-of-drug-use) and the other athletes in the [1988 Summer Olympics](https://en.wikipedia.org/wiki/Athletics_at_the_1988_Summer_Olympics_%E2%80%93_Men%27s_100_metres). Johnson took [Stanozolol](https://en.wikipedia.org/wiki/Stanozolol), an anabolic agent popular among sprinters because of it's ability to gain lean muscle mass.  

**Steroid Cycles:** Generally speaking, anabolic agents are used in the pre-competition or "building" phase to enhance an athlete's training, with a focus on developing size, power, or strength. Starting and stopping steroids use throughout a competitive season is referred to as "cycling" (or a "cycle"). The duration of cycles vary, but generally don't exceed 8 weeks. Sports and athletic events in which size or strength contribute to performance have a potential market for anabolic agents. In fact, coaches and athletes use various resistance training programs and diets to try and stimulate these processes throughout an athletic career. 

These are stored in the string `anabolic_agents` below





```{r anabolic_agents}
anabolic_agents <- c(
  "1-androstenedione",

  "1-testosterone and 1-androstenedione",

  "17β-hydroxymethyl-17α-methyl-18-norandrost-1,4,13-trien-3-one",

  "19-norandrostendione",
  "19-norandrosterone",
  "19-norandrosterone (19-na)",
  "19-norandrosterone & 19-noretiocholanolone",
  "19-norandrosterone and 19-noretiocholanolone",
  "19-norandrosterone and testosterone",
  "19-norandrosterone, epo, furosemide",

  "6a-hydroxyandrostendione",
  "6a-oh-androstenedione & 6b-ohandrosterone",
  "anastrozole",
  
  "androgenic-anabolic steroids",
  "androgenic anabolic steroid",
  "androstendione",
  "androstenedione",
  "androstenedione, ostarine",
  "dehydrochloromethyltestosterone (dhcmt)",
  "dehydroepiandrosterone (“dhea”)",
  "dehydroepiandrosterone (dhea)",

  "boldenone",
  "boldenone and 19-norandrosterone",
  "boldenone metabolite",
  "boldione",
  "carboxy finasteride",
  "dianabol",

  "human growth hormone",
  "human growth hormone (hgh)",
  "human growth hormone (hgh) and testosterone",

  "clenbuterol",
  "clenbuterol, ostarine, and a tamoxifen metabolite",
  "clomiphene",
  "clomiphene and gw1516",
  "clomiphene, letrozole",
  "clomiphene, methylphenidate metabolite",
  "formestane (4-hydroxyandrostendione)",

  "testosterone",
  "testosterone & modafinil",
  "testosterone & recombinant human erythropoietin (r-epo)",
  "testosterone and boldenone",
  "testosterone metabolites (exogenous origin)",
  "testosterone prohormones",
  "testosterone, methandienone (dianabol) and stanzolol",
  "testosterone, nandrolone & hcg",
  "dexamethasone",
  "drostanolone",
  "elevated testosterone-epitestosterone (t/e) ratio",
  "exogenous testosterone",
  "tetrahydrogestrinone (thg)",
  "finasteride",
  "finasteride & 6-oxo-androstenedione",

  "anabolic agent",
  "anabolic agents",

  "ghrp-2",
  "growth hormone-releasing hexapeptide",
  "gw1516",
  "hormone antagonists, hormones and related substance, hcg, and an anabolic agent",
  "insulin-like growth factor-1",
  "methandienone",
  "methandienone (dianabol)",
  "ibutamoren",
  "meldonium",
  "metabolites of stanozolol",
  "methasterone, ostarine, androstatrienedione",
  "methenolone, androst-(2,3)-en-17-one, and ghrp-6",
  "methylclostebol, promagnon",
  "methylprednisolone",
  "methyltestosterone",
  "synthetic testosterone",
  "proplyhexedrine; anabolic steroid",
  "nandrolone metabolites",
  "norandrosterone",
  "stanozolol",
  "stanozolol / methylphenidate",
  "stanozolol metabolites",
  "stanzolol",
  "ostarine",
  "ostarine and 7-keto-dhea",
  "ostarine, arimistane",
  "oxandrolone",
  "oxandrolone and clenbuterol",
  "oxymetholone",
  "prednisone",
  "rad140 and ibutamoren",
  "tamoxifen",
  "te ratio > 4:1 & hcg",
  "trenbolone",
  "trenbolone, epitrenbolone",
  "triamcinolone acetonide",
  "ostarine"
)

anabolic_agents_regex <- paste0("(", str_c(anabolic_agents, collapse = "|"), ")")

str_view_all(substances, anabolic_agents_regex, match = TRUE)
```

Now I can use this regex pattern to assign a new category inside `dplyr::mutate()` and `dplyr::case_when()`.

```{r S1S2_agents}
UsadaSanct <- UsadaSanct %>%
  dplyr::mutate(
    S1S2_agents =
      case_when(
        str_detect(substance_reason, anabolic_agents_regex) ~ substance_reason,
        TRUE ~ NA_character_))

UsadaSanct %>%
  dplyr::select(
    athlete,
    S1S2_agents,
    substance_reason) %>%
  dplyr::filter(!is.na(S1S2_agents)) %>%
  dplyr::arrange(desc(S1S2_agents))
```



## 2. Stimulants & diuretics     

```{r stimulants_diuretics}
stimulants_diuretics <- c(
  "1,3-dimethylbutylamine",
  "1,3-dimethylbutylamine (dmba)",
  # is a stimulant drug structurally related to methylhexanamine where
  # a butyl group replaces the amyl group
  "amphetamine",
  "amphetamine & methamphetamine",
  "amphetamine and methylhexaneamine",
  "adverse cir & amphetamine",
  # a prohibited stimulant
  "acetazolamide",
  # Diuretic - It can treat glaucoma, epilepsy, mountain sickness,
  # and fluid retention (edema).
  "benzoylecgonine",
  "benzoylecgonine, a metabolite of cocaine",
  # It is the corresponding carboxylic acid of cocaine
  "canrenone",
  # a steroidal antimineralocorticoid of the spirolactone group related
  # to spironolactone which is used as a diuretic
  "cathine (norpseudoephedrine)",

  "chlorothiazide & hydrocholorthiazide",

  "ephedrine",

  "hydrochlorothiazide",
  "hydrochlorothiazide & chlorothiazide",
  "hydrochlorothiazide & triamterene",
  "hydrochlorothiazide and chlorothiazide",
  "hydrochlorothiazide and triamterene",
  "hydrochlorothiazide, chlorothiazide",
  "hydrochlorothiazide, chlorothiazide, triamterene and labetalol",
  "hydrochlorothiazide, chlorothiazide, triamterene, 4-hydroxytriamterene",
  "hydrochlorothiazide, triamterene",
  # Diuretic - It can treat high blood pressure and fluid retention (edema).

  "pseudoephdrine",
  "pseudoephedrine",
  "pseuedoephedrine",

  "dimethylbutylamine",
  "dimethylbutylamine (dmba)",

  "l-methamphetamine",

  "furosemide",
  "furosemide and refusal",

  "methylhexanamine",
  "methylhexaneamine",
  "methylhexaneamine & phenpromethamine",

  "methylphenidate",

  "methylphenidate - ritalinic acid",

  "modafinil",

  "zeranol",

  "nikethamide",

  "oxilofrine",

  "phentermine",

  "phentermine, boldenone and nandrolone",

  "tuaminoheptane"
)
```

## 3. Blood doping/Erythropoietin (EPO)  

```{r blood_dope_epo}
blood_dope_epo <- c(
  "blood doping",
  "epo",
  "epo, hgh, cortisone",
  "epo, hgh, testosterone",
  "epo, hgh, testosterone, blood transfusions",
  "epo, hgh, testosterone, cortisone",
  "epo, hgh, testosterone, cortisone, blood transfusions",
  "epo, testosterone, blood transfusions",
  "epo, the testosterone/epitestosterone cream, and thg",
  "erythropoiesis-stimulating agent",
  "erythropoietin (epo)",
  "erythropoietin (epo), human growth hormone (hgh), and testosterone",
  "efaproxiral",
  "human recombinant erythropoietin",
  "recombinant human erythropoietin (repo)",
  "use and possession of epo & evading sample collection",
  "use, attempted use and possession of epo",
  "oxygen-enhancing peptide hormone",
  "r-epo"
)
```

## 4. Cannabinoids  

```{r cannabinoids}
cannabinoids <- c(
  "11-nor-tetrahydrocannabinol-9-carboxylic acid",
  "metabolites of cannabis",
  "metabolites of cannabis or tetrahydrocannabinol (thc)",
  "metabolites of cannabis or tetrahydrocannibinol (thc)",
  "cannabidiol",
  "cannabis",
  "carboxy-tetrahydrocannabinol (thc)",
  "carboxy-thc",
  "thc",
  "thc metabolites",
  "tetrahydrocannabinol",
  "tetrahydrocannabinol acid"
)
```

## 5. Beta blockers  

```{r beta_blockers}
beta_blockers <- c(
  "formoterol",
  "metoprolol",
  "vilanterol",
  "beta-2 agonist",
  "higenamine",
  "higenamine, 1,4‐dimethylpentylamine",
  "salbutamol",
  "salmeterol"
)
```

## 6. Multiple substances     

```{r multiple_substances}
multiple_substances <- c(
  "ostarine, stanozolol metabolites 16β-hydroxystanozolol and 4β-hydroxystanozolol,  hydrochlorothiazide, chlorothiazide",
  "erythropoietin (epo), anabolic steroids, modafinil",
  "amphetamine, and metabolites of drostanolone, mesterolone, and dehydrochloromethyltestosterone (dhcmt)",
  "adverse cir, anastrozole, lgd-4033, & ostarine",
  "adverse cir",
  "anabolic androgenic steroid, rhepo, amphetamine",
  "anabolic steroids, insulin, erythropoietin (epo), growth hormone and modafinil",
  "anabolic androgenic steroid / adverse cir",
  "thc and amphetamine",
  "methylphenidate, anabolic agents/metabolites",
  "modafanil - exogenous anabolic androgenic steroids",
  "tamoxifen; boldenone, methandienone, drostanolone, and clenbuterol",
  "epo, testosterone, blood transfusions and corticosteroids, trafficking, administration to others",
  "ritalinic acid (ritalin)",
  "use, attempted use and possession of epo and hgh",
  "raloxifene, ostarine, ibutamoren, gw1516 sulfone, rad140, lgd4033, andarine",
  "strychnine",
  "tetrahydrogestrinone (thg) and modafinil"
)
```

## 7. Pain killers      

```{r pain_killers}
pain_killers <- c(
  "fentanyl",
  "oxycodone",
  "noroxycodone"
)
```

## 8. Masking agents  

```{r masking_agents}
# what other masksing agents?
masking_agents <- c("probenecid")
```

## 9. Glycerol (soon to change) 

```{r glycerol}
# ban is about to be lifted
glycerol <- c("glycerol")
```


## 10. Non-substance sanctions

```{r}
non_drug_sanctions <- c(
  "test evasion",
  "intravenous infusion",
  "test refusal",
  "trafficking and administering prohibitied substances",
  "violating period of ineligibility",
  "3 missed tests",
  "3 whereabouts failures",
  "failed to appear",
  "failure to appear",
  "failure to appear for test",
  "failure to appear for testing",
  "manipulation of forms",
  "missed test violation",
  "missed tests",
  "non-analytical",
  "non-analytical positive",
  "non-analytical, ghrfs",
  "non-analytical, possession of prohibited peptides",
  "non-analytical; possession, trafficking, and administration of prohibited substances",
  "possession, trafficking, and administration of prohibited substances",
  "possession, trafficking, and administration of prohibited substances and methods",
  "reduced sanction",
  "sanction restarted for violation",
  "refusal to submit to doping control",
  "refusal to submit to sample collection",
  "refusal to test",
  "refusal/non-analytical positive",
  "tampering"
)
```




